Online {\IterComp}

### 2017-09-15 abstract.tex
changes to abstract

    Iterative or adaptive compilation can dramatically improve the performance of a program by searching for the best optimization
    settings. However, it relies upon repeatedly evaluating and comparing optimizations off-line on predefined
    representative inputs. If the developers fail to guess what is a representative input or usage patterns change over time, 
    then the wrong optimizations may be selected. 
    Searching the optimization space with actual user inputs, during deployment, has previously seemed infeasible. 
    Program side effects, interactivity and difficulties capturing the system state mean that any given input can 
    only be executed once. Since each input may entail a different amount of work, the search cannot use runtime to compare 
    optimizations, and there is no baseline against which to define speedup.
    Unless we find a way to compare optimizations even when inputs are executed only once, we will be unable to use real user 
    input for iterative compilation, users will not have the best optimized programs, and performance and energy efficiency will suffer.

    We present metrics and techniques which, for the first time,  enable the use of true {\em online iterative compilation}.
    We compare optimizations by a new work efficiency metric, which requires only a single execution with any one input.
    We develop an instrumented, low overhead, mechanism for gathering the required data, using optimal probe placement in the source
    code. We show that overhead can be further reduced by removing high-cost probes whose absence introduces only small errors to the 
    work metric, without significantly affecting the outcome of the iterative compilation. 
    We give two variants for probe removal optimization: the first offers strong
    guarantees about the maximum possible error, while the second targets the average, whole program error.

    Our online iterative compilation method yields programs that get 80\% of the speedup achievable by an offline oracle, 
    but without executing any input more than once, which the oracle cannot do. 
    Through probe removal we bring the instrumentation overhead down to only 4\% on average, making our
    approach suitable for regular use. We show practical online iterative compilation for the first time, 
    optimizing programs according to the range of inputs users actually present.

\end{abstract}

>>>>

    Iterative compilation can dramatically improve the performance of a program by searching for the best optimization.
    However, it relies upon repeatedly evaluating optimizations on predefined representative inputs.
    If the developers fail to guess what is a representative input or usage patterns change over time,
    then the wrong optimizations may be selected.
    Searching the optimization space with actual user inputs, during deployment, has previously seemed infeasible.
    Program side effects and difficulties capturing the system state mean that any given input can
    only be executed once. %Since each input may entail a different amount of work, the search cannot use runtime to compare
    %optimizations, and there is no baseline against which to define speedup.
    Unless we find a way to compare optimizations even when inputs are executed only once, we will be unable to use real user
    input for iterative compilation, users will not have the best optimized programs, and performance will suffer.

    We present metrics and techniques which, for the first time,  enable the use of true {\em online iterative compilation}.
    We compare optimizations by a new work efficiency metric, which requires only a single execution with any one input.
    We develop an instrumented, low overhead, mechanism for gathering the required data, using optimal probe placement in the source
    code. We show that overhead can be further reduced by removing high-cost probes whose absence introduces only small errors to the
    work metric, without significantly affecting the outcome of the iterative compilation.
    We give two variants for probe removal optimization: the first offers strong
    guarantees about the maximum possible error, while the second targets the average whole program error.

    Our online iterative compilation method yields programs that get 80\% of the speedup achievable by an offline oracle,
    but without executing any input more than once, which the oracle cannot do.
    Through probe removal we bring the instrumentation overhead down to only 5\% on average, making our
    approach suitable for real-world scenarios. %We show practical online iterative compilation for the first time,
    %optimizing programs according to the range of inputs users actually present.
\end{abstract}

### 2017-09-15 results.tex
changes to abstract

A few of the benchmark present abnormal regressions in their execution time.
This increase in the profiling overhead is counter-intuitive because the
relaxation algorithms only reduces the number of instrumented probes,
without changing their placement.
We suggest some reasons for this abnormal regressions.
By analysing these cases, we noticed that several of the frequently executed
probes could not be removed by the relaxation.
Moreover, another important aspect to point out is that we also observed
differences among the final CFGs after applying the \texttt{-O3} optimization.
Although relaxation does not directly alter the CFG, it is worth mentioning that
many optimization passes make use of LLVM's built-in target-based cost model,
which means that adding or removing instructions from basic blocks may affect
decisions during transformations of the code.
For example, some of the optimizations that make use of this built-in cost model
and also alter the CFG are: loop-unrolling, function inlining, and
simplifications of the CFG.
These changes in the program's CFGs may affect other compiler optimizations and
also possibly in hardware performance, for example affecting the instruction
caching.

Furthermore, these improvements can be obtained while incurring very little dynamic error, as shown in Figure~\ref{fig:error-O3}.
This result confirms the overly conservative aspect of the relaxation algorithm applied per DAG,
while the WPO relaxation is significantly less conservative.
Although the error bound is not guaranteed by the WPO relaxation,
its dynamic error was always below the 5\% thresholds due to the use
of perfect profiling information.

@@

The comparison between Oracle-RM and Oracle-PP is crucial for validating the
effectiveness of the work efficiency metric for guiding online {\itercomp},
while the other configurations demonstrate the viability of applying online
{\itercomp} in real-world scenarios.
For the relaxation strategies, we evaluate only those with a 5\% threshold, since
they offered better overhead reductions for little dynamic error in the measurement.

>>>>

A few of the benchmarks present abnormal regressions in their execution time.
This increase in the profiling overhead is counter-intuitive because the
relaxation algorithms only reduces the number of instrumented probes,
without changing their placement.
We suggest some reasons for these abnormal regressions.
By analyzing these cases, we noticed that several of the frequently executed
probes could not be removed by the relaxation strategies with small thresholds.
Another important aspect to point out is that we also observed
differences among the final CFGs after applying the \texttt{-O3} optimization.
Although relaxation does not directly alter the CFG, it is worth mentioning that
many optimization passes make use of LLVM's built-in target-based cost model,
which means that adding or removing instructions from basic blocks may affect
decisions during transformations of the code.
For example, some of the optimizations that make use of this built-in cost model
and also alter the CFG are: loop-unrolling, function inlining, and
simplifications of the CFG.
These changes in the program's CFGs may affect other compiler optimizations and
also possibly in hardware performance, for example affecting the instruction
caching.

Finally, as shown in Figure~\ref{fig:error-O3}, these improvements can be
obtained while incurring very little dynamic error.
This result confirms the overly conservative aspect of the relaxation algorithm
applied per DAG, while the WPO relaxation is significantly less conservative.
Although the error bound is not guaranteed by the WPO relaxation,
its dynamic error was always below the 5\% thresholds due to the use
of perfect profiling information.

@@

The comparison between Oracle-RM and Oracle-PP is crucial for validating the
effectiveness of the work efficiency metric in guiding online {\itercomp},
while the other configurations demonstrate the viability of applying online
{\itercomp} in real-world scenarios.
For the relaxation strategies, we evaluate only those with a 5\% threshold, since
they offered better overhead reductions for little dynamic errors in the measurement.
